Lexicalized Domain Representations for Neural Machine Translation

### 2019-05-21 2019/emnlp/sparse.tex
rewrite algorithm

Multi-domain adaptation is conceptually close to multilingual MT, or more generally to multi-task learning \cite{Caruana97multitask}, and can be approached in a number of ways. We adapt here ideas of \citet{Daume07frustratingly} to neural MT, in an attempt to confine the differences between domains at the lexical level, while the rest of the parameters is shared between domains. Our main hypothesis is that most of the differences between domains take place at the lexical level, due to cross-domain polysemy. By contrast, the deeper layers, which arguably model more abstract linguistic phenomena, are made shareable across domains.
To this end, we design word embeddings made of a generic as well as several domain-specific regions. We perform experiments with three domains, two neural architectures and two language pairs and find that our technique can effectively build multi-domain NMT, outperforming several baselines. Our contribution is thus as follows:
\begin{itemize}
\item we implement the ideas of \cite{Daume07frustratingly} for two NMT architectures;
\item we provide experimental evidence that suggest the effectiveness of this technique;
\item our analysis of embeddings seems to confort our assumption that their variation across domain actually reflects a variation of senses.
\end{itemize}
\fyDone{can we train in random order ? can we get away with catastrophic forgetting ?}
\fyDone{how to analyze the embeddings ? how can we test or claim ?}

\section{Lexicalize domain representations\label{sec:lexicalized_embeddings}}
\fyDone{Use meaningful titles throughout}

@@

\begin{algorithm}[h]
\caption{Multi-domain Training}
\label{alg:multidomain}
\begin{algorithmic}[1]
\REQUIRE {Corpora $C_i, i\in [1,..,d]$ for $d$ domains, Batch size $B$, Optimization algorithm $\operatorname{Opt}$}
\REPEAT 
\STATE{Randomly pick $i \in [1,..,d]$ w.r.t the multinomial distribution $[\frac{|C_i|}{\displaystyle{\mathop{\sum}_{i\in [1,..,d]}|C_i|}}]$.}
\STATE{Randomly pick $B$ sentences from $C_i$.}
\STATE{Activate only generic region to create generic embedding batch, denoted $W_g$.}
\STATE{Pass the generic embedding batch to a dense layer.}
\STATE{Compute gradient of $\theta_s$, $\frac{\partial L}{\partial \theta_s}$ using $W_g$.}
\STATE{Activate domain region and generic region to create domain embedding batch $W_i$}
\STATE{Pass $W_i$ to a dense layer}
\STATE{Compute gradient of domain-specific parameters $\theta_i$, $\frac{\partial L}{\partial \theta_i}$ using $W_d$.}
\STATE{Apply learning rules using computed gradients: 
$$ \theta_s = \theta_s - \operatorname{Opt}(\frac{\partial L}{\partial \theta_s}(W_g))$$ 
$$ \theta_i = \theta_i - \operatorname{Opt} (\frac{\partial L}{\partial \theta_i}(W_i))$$}
\UNTIL{convergence}
\end{algorithmic}
\end{algorithm}

@@

% We also take into account the problem of overfitting that could occur for small corpora. Indeed, real data are not always balanced for every domain. For instance, in our experiments, the size of corpora are very different where EPPS is largest with approximately 2 millions sentences, EMEA contains about 1 million sentences and ECB is the smallest with only 200 000 sentences for English French and 100 000 for English-German. If we train each corpora with same number of batches, ECB will be trained too oftenly, that easily leads to overfitting. Therefore, we select batches from each corpora with frequencies proportional to its size.
The batch selection procedure (step~2 of algorithm~\ref{alg:multidomain}) ensures that each domain is represented according to its frequency in the training data. In our experiments, this means that sentences from the Europarl domain will be selected more frequently that the two other domains. In an attempt to build a more balanced training procedure, we have also experimented with an alternative selection procedure, where $i$ is selected according to the multinomial distribution $[\frac{\sqrt{|C_i|}}{\displaystyle{\mathop{\sum}_{i\in [1,..,d]}\sqrt{|C_i|}}}]$.
\fyTodo{Refs on this ? or contrast ?}

@@

For these test sets, our proposal (\texttt{LDR}$_{oracle}$) is consistently better that the mix-domain strategy, with gains that range from very large (for EMEA and ECB) to unsignificant (for EPPS in most conditions). This means that our architecture is somehow able to compensate for the data unbalance and to raise the performance of the multi-domain system close to the best (fine-tuned) system in each domain. We even observe rare cases where the \texttt{LDR}$_{oracle}$ system outperforms fine-tuning (eg.\ Transformer en:ge in the EMEA domain). \texttt{LDR}$_{oracle}$ is also better than Domain Control in three conditions out of four, \texttt{DC} being seemingly a better choice for the RNN than for the Transformer architecture. The comparison with the \texttt{WDCMT} model is less favourable for our RNN systems, which are arguably much simpler.\footnote{The direct comparison with WDCMT is difficult as the implementation differs in many ways from our RNN: a different framework, different cell types,
  % (LSTMs vs. conditional GRUs),
  etc.} \fyDone{More on this.}
As expected, ignoring the domain label yields a drop in performance for the two fallback strategies considered in this study. The first, \texttt{LDR}$_{generic}$, only uses the generic part of the embeddings. The second, \texttt{LDR}$_{pred}$, resorts to automatically predicted domain labels\footnote{Our domain classifier uses a bi-LSTM RNN encoder, followed by a simple softmax layer. Its precision on a development set are higher than 95\%.}
\fyTodo{Explain how} Note that this decrease is however quite limited (and hardly significant for \texttt{LDR}$_{pred}$), showing that our architecture is quite robust to noisy labels. Even in the worst case scenario where all domain tags are intently wrong (\texttt{LDR}$_{wrong}$), we see that the generic part still ensures a satisfying level of performance. A last contrast is with \texttt{LDR}$_{oracle}^{0.5}$ where we change the distribution of training sentence to increase the number of ECB samples: as a result, we see a small decrease for EMEA and EPPS, and a large boost for ECB. This shows that our technique can be used in conjunction to other well known strategies for performing domain adaptation.

@@

By analogy with the proposal of \citet{Johnson17google} for multilingual NMT, \citet{Kobus17domaincontrol} and \citet{Chu18multilingual} separately propose to extend the representation of the source text with a domain tag. Our model also modifies input representations, but allows each source word to have its own domain-specific representation, thereby improving training of the shared parts of the network.\fyTodo{Talk about target embeddings somewhere.}

>>>>

Multi-domain adaptation is conceptually close to multilingual MT, or more generally to multi-task learning \cite{Caruana97multitask}, and can be approached in a number of ways. We adapt here ideas of \citet{Daume07frustratingly} to neural MT.
% in an attempt to confine the differences between domains at the lexical level, while the rest of the parameters is shared between domains.
Our main hypothesis is that domains mostly differ at the lexical level, due to cross-domain polysemy, which motivates domain specific embeddings. By contrast, the deeper layers, which arguably model more abstract linguistic phenomena, are made shareable across domains. To this end, we design word embeddings containing a generic and several domain-specific regions. We experiment with three domains, two neural architectures and two language pairs and find that our technique yields effective multi-domain NMTs, outperforming several baselines. Our contribution is thus as follows:
\begin{itemize}
\item we implement the ideas of \cite{Daume07frustratingly} for two NMT architectures;
\item we provide experimental evidence that suggest the effectiveness of this technique;
\item our analysis of embeddings seems to confort our assumption that their variation across domain actually reflects a variation of senses.
\end{itemize}
\fyDone{can we train in random order ? can we get away with catastrophic forgetting ?}
\fyDone{how to analyze the embeddings ? how can we test or claim ?}

\section{Lexicalized domain representations\label{sec:lexicalized_embeddings}}
\fyDone{Use meaningful titles throughout}

@@

\begin{algorithm}[h]
\caption{Multi-domain Training}
\label{alg:multidomain}
\begin{algorithmic}[1]
\REQUIRE {Corpora $C_i, i\in [1,..,d]$ for $d$ domains, Batch size $B$}%, Optimization algorithm $\operatorname{Opt}$}
\REPEAT 
\STATE{Randomly pick $i \in [1,..,d]$ w.r.t the multinomial distribution $[\frac{|C_i|}{\sum_{i\in [1,..,d]}|C_i|}]$.}
\STATE{Randomly pick $B$ sentences from $C_i$.}
\STATE{Activate only generic region to create generic batch, denoted $W_g$.}
% \STATE{Pass the generic embedding batch to a dense layer.}
\STATE{Compute gradient of $\theta_s$, $\frac{\partial L}{\partial \theta_s}$ using $W_g$.}
\STATE{Activate domain-specific and generic regions to create domain-specific batch $W_i$}
% \STATE{Pass $W_i$ to a dense layer}
\STATE{Compute gradient of domain-specific parameters $\theta_i$, $\frac{\partial L}{\partial \theta_i}$ using $W_d$.}
\STATE{Update parameters $\theta_s$ using $\frac{\partial L}{\partial \theta_s}(W_g)$ and $\theta_i$ using $\frac{\partial L}{\partial \theta_i}(W_i)$}
% $$ \theta_s = \theta_s - \operatorname{Opt}(\frac{\partial L}{\partial \theta_s}(W_g))$$ 
% $$ \theta_i = \theta_i - \operatorname{Opt} (\frac{\partial L}{\partial \theta_i}(W_i))$$}
\UNTIL{convergence}
\end{algorithmic}
\end{algorithm}

@@

% We also take into account the problem of overfitting that could occur for small corpora. Indeed, real data are not always balanced for every domain. For instance, in our experiments, the size of corpora are very different where EPPS is largest with approximately 2 millions sentences, EMEA contains about 1 million sentences and ECB is the smallest with only 200 000 sentences for English French and 100 000 for English-German. If we train each corpora with same number of batches, ECB will be trained too oftenly, that easily leads to overfitting. Therefore, we select batches from each corpora with frequencies proportional to its size.
The batch selection procedure (step~2 of algorithm~\ref{alg:multidomain}) ensures that each domain is represented according to its frequency in the training data. In our experiments, this means that sentences from the Europarl domain will be selected more frequently that the two other domains. In an attempt to build a more balanced training sample, we also consider an alternative selection procedure, where $i$ is selected according to the distribution $[\frac{\sqrt{|C_i|}}{\sum_{i\in [1,..,d]}\sqrt{|C_i|}}]$.
% \fyTodo{Refs on this ? or contrast ?}

@@

For these test sets, our proposal (\texttt{LDR}$_{oracle}$) is consistently better that the mix-domain strategy, with gains that range from very large (for EMEA and ECB) to unsignificant (for EPPS in most conditions). This means that our architecture is somehow able to compensate for the data unbalance and to raise the performance of the multi-domain system close to the best (fine-tuned) system in each domain. We even observe rare cases where the \texttt{LDR}$_{oracle}$ system outperforms fine-tuning (eg.\ Transformer en:ge in the EMEA domain). \texttt{LDR}$_{oracle}$ is also better than Domain Control in three conditions out of four, \texttt{DC} being seemingly a better choice for the RNN than for the Transformer architecture. The comparison with the \texttt{WDCMT} model is less favourable for our RNN systems, which are arguably much simpler.\footnote{The direct comparison with WDCMT is difficult as the implementation differs in many ways from our RNN: a different framework, different cell types,
  % (LSTMs vs. conditional GRUs),
  etc.} \fyDone{More on this.}
As expected, ignoring the domain label yields a drop in performance for the two fallback strategies considered in this study. The first, \texttt{LDR}$_{generic}$, only uses the generic part of the embeddings. The second, \texttt{LDR}$_{pred}$, resorts to automatically predicted domain labels\footnote{Our domain classifier uses a bi-LSTM RNN encoder, followed by a simple softmax layer. Its precision on a development set are higher than 95\%.}
\fyDone{Explain how} Note that this decrease is however quite limited (and hardly significant for \texttt{LDR}$_{pred}$), showing that our architecture is quite robust to noisy labels. Even in the worst case scenario where all domain tags are intently wrong (\texttt{LDR}$_{wrong}$), we see that the generic part still ensures a satisfying level of performance. A last contrast is with \texttt{LDR}$_{oracle}^{0.5}$ where we change the distribution of training sentence to increase the number of ECB samples: as a result, we see a small decrease for EMEA and EPPS, and a large boost for ECB. This shows that our technique can be used in conjunction to other well known strategies for performing domain adaptation.

@@

By analogy with the proposal of \citet{Johnson17google} for multilingual NMT, \citet{Kobus17domaincontrol} and \citet{Chu18multilingual} separately propose to extend the representation of the source text with a domain tag. Our model also modifies input representations, but allows each source word to have a domain-specific representation, thereby improving training of the shared parts of the network.\fyDone{Talk about target embeddings somewhere.}
